Tail associations in ecological variables and their impact on extinction risk

### 2020-01-20 Response.txt

3.      The amount of analyses for metapopulations and age-structured populations seems quite unbalanced. Why is this so? Obviously, the paper would greatly increase its value by including more analyses for stage-structured populations, but then the paper may become too long to fit in the required length for Ecosphere. Would it be more beneficial to focus only on metapopulations in this paper and expand on the stage-structured population part in a separate paper to more deeply examine Tuljapurkar’s approximation?

>>>Thank you for asking these questions. First, this is an initial attempt to understand the impact of asymmetric tail associations on extinction risk, and as such we intentionally used a simple approach for a rapid assessment of whether the basic ideas have merit. Based on our results, we believe this could turn into a broad topic, which many researchers should investigate. Our intention was to provide an initial exploration of the mechanisms rapidly, and in so doing, hopefully to open what could be a practically important topic (for conservation) to consideration by other researchers, as quickly and broadly as possible. With regard to extensive mathematical analyses (e.g. using Tuljapurkar's approaximation) of stochastic matrix models, which the referee suggests, we are journeymen only, not true experts. Mathematical analysis of stochastic matrix models is a deep field. So we believe carving our matrix model results out of the current paper and then undertaking a more systematic explortion ourselves, for possible later publication, does not get these potentially practically important ideas out there as fast and risks the possibility that we will not be able to do the analyses as well as some of our readers.
  Second, for stochastic matrix models, our initial explorations suggested that probably the effects of tail association in that context are minor, so we went no further. In contrast, for metapopulation models, our initial explorations (with the Ricker model) showed a likely strong effect, so we went further with additional models. So what the referee perceives as imbalance is actually just a rational attempt to explore promising ideas further and to simply identify other ideas as less promising, to help future researchers choose what to work on.  
  Finally, we do not see as much of an imbalance as the referee suggests. To us, the two components are treated with slightly different throroughness, but not markedly so, and for good reasons (see above). Our matrix modelling results are certainly enough to make the points we wanted to make, i.e., an initial assessment that the influence of tail associations on extinction risk in this context seems likely to be minor.

@@

4.      The order of appearance of the metapopulation problem and the age-structured population problem keeps switching throughout the paper. For readability, the order should be consistent.

>>>Thanks for raising this point. We have edited to enforce consistency of this ordering throughout the manuscript.

@@

5.      Whether variables have tail associations can be altered by transformation; e.g., variables may have no tail association in one way but show any of the tail associations discussed in this paper in another. How should researchers who want to use the approached introduced in this paper deal with this? Should they look for a transformation that can give one of such tail associations as long as it is ecologically meaningful in a given system? I suppose so, but this is not mentioned in the paper.

>>>The referee is mistaken here, transformation does not alter tail associations as we have defined them based on data ranks. This is a standard way of defining tail associations and copula approaches to the study of dependencies, justified by the Ghosh et al paper (which, by the way, has now been formally accepted at Advances in Ecological Research, and is in press) and by Genest & Favre, 2007, J Hydol Eng, "Everything you always wanted to know about copula modelling but were afraid to ask". We have added an explanation of this fact in the Discussion, and made some related points here that will help readers in future applications of our ideas.

***DAN: Don't forget to add the explanation.

@@

Also, the interpretation that certain variables are “bad” or “good” depends on how it is formulated or transformed. It could be that, if a “bad” variable is associated with a “good” variable in their right tails, for example, their effects cancel each other and leave no overall effect, which would be also an interesting result. This may be related to the discussion about negative associations between Line 411 and 423, but it does not clearly come across. Maybe these are discussed in the Ghosh et al. paper in BioRxiv, but it is not yet published in a peer reviewed journal. These aspects should be worth mentioning in this paper for clarification of the method.

>>>The Ghosh et al paper is now formally accepted. The version on the BioRxiv is essentially the same as the final, accepted version. In spite of thinking for a quite a long time about what the referee may have meant by the above comment, we confess we are not sure, so we do not know what to do about this comment. In our model formulations, variables are never "good" or "bad" but are instead assumed to fluctuate, taking values in some years which are "good" for the population (by which we mean they promote increased reproduction or increased survival in that location, in that year), and values in other years which are "bad" (in the opposite sense). Aside from an apparent desire to have us make some additional point (I'm afriad we cannot understand what point), the referee does not appear to be criticizing anything we actually do say. So we suggest proceeding with this text as it is. As we said, the Ghosh et al paper is now accepted, and is already available online on the BioRxiv, and many important related points are made there. So readers looking to base some of their own research on the points made in the current manuscript can also use the Ghosh et al paper as an additional background resource before getting started.

@@

6.      Figure 5 presents cases with d = 0 only and in the appendix different values of d are considered only for global dispersal. Would the results qualitatively hold when different values of d for local dispersal is considered?

>>>If we understand correctly what the referee is saying here, he/she seems to be mistaken. Figs S2-S7 show results for each model for a whole range of d, for local dispersal. These figures are referenced at the end of the second-to-last paragraph of Results, and in the last paragraph of Results. It can be determined through examination of Figs S2-S13 that yes, results will qualitatively hold for nonzero d for the types of dispersal we considered (local and global). A statement to this effect has been added to the last paragraph of Results.

@@

7.      In general, the text in certain parts can be much tightened for readability and clarity.

>>>In the course of doing other edits we went through the entire text again and tightened where possible.

***DAN: make sure to do this

Minor comment
•       Line 97-105. It is too abstract to follow in the current form until the reader read the last sentence.

>>>We reworded the paragraph.

•       Line 100, insert “same” in between “the” and “variable” for clarity.

>>>Done.

@@

•       As I mentioned above, Q1 and Q2 appear in an opposite order from what I expected from the introduction so far.

>>>We have fixed this ordering issue throughout the manuscript, see our response to point 4 above.

@@

•       Line 299. Because simulation studies can increase the number of replicates relatively easily and usually use a large number of replicates, statistical significance does not mean much in most of cases, including this one.

>>>Thank you, we take the referee's point here and have eliminated the statement about significance.

@@

•       Line 357-9. Rewrite the sentence. The paper did not show “how such (climatic) changes modify the asymmetry of tail association of weather variables measured in different places.” It showed only whether the latter part of the sentence happens.

>>>The sentence the referee mentions was the topic sentence of the paragraph. The first part of that sentence was what the argument of the paragraph is about. So yes, we did do this, it was accomplished not in the Results of the paper (which is what the referee seems to have expected), but instead was accomplished at this point in the manuscript, in this very paragraph. We apologize, however, that this was unclear! So we reworded to make it clearer. We also worked on this whole paragraph, and the following one, to make them clearer.

@@

•       In the second paragraph in Discussion, X’s are used to refer to generic random variables, but it appeared rather abruptly as the letter was not used to denote variables in the paper up to that point in the main text. It would be helpful to give the definition parenthetically in the text. Do i and j refer to different weather variables, to different ecological variable, or also to different patches?

>>>We apologize for this omission! We have now added the definition. We have also clarified the rest of this paragraph.

@@

•       The paragraph starting from Line 372 is very long and winding. This needs to be tightened substantially (to 2/3 to ½ of the current length, which is 1.5 pages).

>>>We split the paragraph into two and shortened, as the referee requesrs.

@@

•       Line 459-460. Please give a brief definition of “the geography of synchrony,” and clarify what “synchrony was found to have a prominent geography…” means here.

>>>Done, and the clarity is improved. Thanks.

Please see also the attached file.

***DAN: Shyamolina, did you look carefully through the attached file mentioned above to see if it had any additional comments, not represented above?

### 2020-01-20 MainText.Rmd
getting close to being done addressing Ecosphere referee comments

<!--
Alternative titles
1) Tail associations in environmental variables and extinction risk
2) The influence of correlations in the extremes on extinction risk
3) Tail associations in environmental variables and their impact on extinction risk
4) Asymmetric tail association in extreme climatic events affects population-extinction risk
-->

@@

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between values of an environmental 
variable measured in different locations, i.e., associations that 
occur between values in the left or right tails of the distributions 
of the variable that occur in the locations. 
Two positively associated variables can be principally "left-tail associated" (i.e., more 
correlated when they take low values than when they take high values) or "right-tail associated"
(more correlated when they take high than low values), even with the same
overall correlation coefficient in both cases. We tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of an environmental variable reduce the growth 
rates of local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental "catastrophes" are
spatially synchronized, presumably reducing the potential for rescue effects.
We also tested, in the context of non-spatial stage-structured matrix models, 
whether tail associations between stage-specific vital rates may
influence extinction risk. 
We showed that density dependence interacts with the
propensity for tail associations to influence metapopulation extinction risk: for 
population models
showing undercompensatory density dependence, left-tail associations in environmental 
variables typically accentuated
and right-tail associations mitigated extinction risk; whereas the 
reverse was usually true for models showing overcompensatory density dependence.
For the non-spatial, stage-structured models we considered, left-tail associations 
between vital rates did accentuate extinction risk compared to right-tail
associations, but the effect was small. 
Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence population extinction risk. 
Our modelling results 
provide an initial demonstration of a new mechanism influencing extinction risks,
and, in our view, should help motivate more comprehensive
study of the mechanism and its importance for real populations in future work.

@@

<!--ECEs probably produce tail associations-->

<!--***DAN wrote this-->
<!--There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, suppose a single ECE is associated
with spatially widespread, large values 
of some climatic variable. Also suppose that small values of the
variable are associated, instead, with local weather, which can differ from location
to location, rather than with ECEs. The result could
be stronger right-tail association, through time, between measurements of the climatic variable
made in different locations.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas. But periods with less rain may experience
scattered showers only, which may be much less correlated across locales.-->

<!--***Shya: I rephrased the following para-->
There are reasons to hypothesize that different locations when influenced by single ECEs may 
show asymmetric tail associations through time. This asymmetry arises because a single 
ECE is associated with spatially widespread, large values 
of some climatic variable whereas small values of the same
variable are associated, instead, with local weather, which can differ from location
to location, rather than with ECEs. For example, rainstorms which deposit 
exceptionally large amounts of rain may do so over large areas. But periods with 
less rain may experience scattered showers only, which may be much 
less correlated across locales. The result could
be stronger right-tail association, through time, between measurements of rainfall
occurred in different locations.

@@

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2000] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Though differences were significant (confidence
bands in Fig. \ref{fig_extrisk_agestr}, which are very narrow, did not overlap),
they were also small, corresponding to only a few percentage points of additional extinction
risk after 25 years. A potential reason for this outcome is elaborated in the Discussion.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} 
and the $r_c$ values of Table \ref{tab_model_summary} reveals that, for the Verhulst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
right-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Verhulst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the opposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across all or most habitat patches. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below).

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechanisms we explored in this paper. 
Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. However, documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE, given that one of them is so
influenced, should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based methods (these studies used Pearson correlations).
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, including how such 
associations may be changing. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be an appropriate and important quantity to study 
for many applications. For instance, in cases for which $X_i$ has a linear 
or approximately linear influence on some aspect of the ecosystem
at location $i$, Pearson correlations of the ecological impacts at locations 
$i$ and $j$, which may be of fundamental interest, will relate to 
Pearson correlations between $X_i$ and $X_j$. But in many cases, the
ecological impact of $X_i$ and $X_j$ will be a nonlinear or unknown monotonic function of 
the values of $X_i$ and $X_j$. So studying rank-based correlations
may be more informative, ecologically as well as providing the mathematical advantages
listed above.

<!--future ideas start here-->
Our results suggest some new lines of research on the 
impact of tail associations on growth rates and 
extinction risks. First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with,
for instance, $r_2$, whereas lower values of $r_1$ may not be strongly
associated with $r_2$ because reductions/increases in $r_1$ can
then be traded off against increases/reductions in either 
$r_2$ or $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we have observed by studying positively associated vital rates.

>>>>

@@

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between distinct population vital rates,
or between values of an environmental 
variable measured in different locations. "Tail associations" between two variables are associations that 
occur between values in the left or right tails of the distributions 
of the variables. 
Two positively associated variables can be principally "left-tail associated" (i.e., more 
correlated when they take low values than when they take high values) or "right-tail associated"
(more correlated when they take high than low values), even with the same
overall correlation coefficient in both cases. 
We tested, in the context of non-spatial stage-structured matrix models, 
whether tail associations between stage-specific vital rates may
influence extinction risk.
We also tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of an environmental variable reduce the growth 
rates of local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental "catastrophes" are
spatially synchronized, presumably reducing the potential for rescue effects.
For the non-spatial, stage-structured models we considered, left-tail associations 
between vital rates did accentuate extinction risk compared to right-tail
associations, but the effect was small. 
In contrast, we showed that density dependence interacts with tail associations 
to influence metapopulation extinction risk substantially: for 
population models
showing undercompensatory density dependence, left-tail associations in environmental 
variables often strongly accentuated
and right-tail associations mitigated extinction risk; whereas the 
reverse was usually true for models showing overcompensatory density dependence.
Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence population extinction risk. 
Our modelling results 
provide an initial demonstration of a new mechanism influencing extinction risks,
and, in our view, should help motivate more comprehensive
study of the mechanism and its importance for real populations in future work.

@@

Extinction risk is very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2000] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates in a population 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different, unrelated factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. The idea, here, is based on the reasonable assumption 
that ECEs are both more intense and spatially more extensive than non-extreme 
climatic events. For clarity, we explain the idea using the example case
of extreme rain events. Suppose a single extreme rain event is associated
with spatially widespread large precipitation values, i.e., not only does it rain a lot
in each locale, but it does so for many locales across a large area. Also suppose that smaller local
precipitation measurements are associated, instead, with local, non-extreme weather 
which can differ from location to location, rather than with a single extreme rain event.
The result could be stronger right-tail association, through time, between measurements of precipitation
made in different locations: large precipation measurements are coherent across space, since 
they are all attributable to a single event; whereas small values are less spatially correlated
because they come from a multiplicity of distinct local events.

@@

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Differences were small, corresponding 
only to a few percentage points of additional extinction risk after 25 years. A 
potential reason for this outcome is elaborated in the Discussion.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} 
and the $r_c$ values of Table \ref{tab_model_summary} reveals that, for the Verhulst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
right-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Verhulst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$. Inspection of Figs
\ref{SM-fig_ricker_risk_local_disp_scl_1}-\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}
shows that a qualitatively similar conclusion holds for nonzero $d$, for both local and global
dispersal.

@@

<!--summary results:-->
Our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below). 
In contrast, our metapopulation results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the opposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across all or most habitat patches.

<!--future ideas start here-->
Our initial exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with,
for instance, $r_2$, whereas lower values of $r_1$ may not be strongly
associated with $r_2$ because reductions/increases in $r_1$ can
then be traded off against increases/reductions in either 
$r_2$ or $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

<!--Tuljapurkar approximation-->
A second topic of research that our work suggests has to do with Tuljapurkar's
well-known approximation of the long-term stochastic growth rate [@tulja1982; @tulja1990].
It may be possible to obtain a better understanding, using Tuljapurkar's
approximation, of our result that asymmetries of tail association
may have limited influence on the extinction risk of 
stage-structured, non-spatial models.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extinction risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations. It seems potentially worth exploring the accuracy of 
Tuljapurkar's approximation under scenarios of tail association in the vital rates
and relating results to the effects of tail associations on the long-term
stochastic growth rate and extinction risks.
We note that the above reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

<!--linking back to ECEs - this para about the metapop case-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We now provide an  
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places. Such modification may thereby influence metapopulation
extinction risk through the mechanisms explored earlier in this paper. 
Let $X_i(t)$ and $X_j(t)$ denote the values of some weather variable as measured in fixed 
locations $i$ and $j$ at time $t$. Although changes in 
the correlation, through time, between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE, given that one of them is so
influenced, should be increasing. That is, since ECEs are spatially bigger, it is now
more likely that a single weather event
influences both locations $i$ and $j$ given that it influences one of them. 
This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored. 
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, in
mean, variance, and possibly other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change, which is instead a change 
in marginal distributions. Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based methods.
